---
title: "Sham vs TAC_RNA-Seq_differential_expression_analysis"
output:
  html_document:
    toc: yes
    toc_float: yes
    code_folding: hide
    fig_width: 10
    fig_height: 8
  pdf_document:
    toc: yes
theme: cosmo
date: "2025-10-18"
---

```{r,packages}


suppressPackageStartupMessages({
library("EnhancedVolcano")
library("limma")
library("biomaRt")
library("annotables")
library("tximport")
library("apeglm")
library("eulerr")
library("DESeq2")
library("HGNChelper")
library("tictoc")
library("DESeq2")
library("kableExtra")
library("beeswarm")
library("missMethyl")
library("gridExtra")
library("png")
library("metafor")
library("ggplot2")
library("purrr")
library("metafor")
library("AnnotationDbi")
library("readxl")
library("ggplot2")
library("tidyverse")
library("magrittr")
library("readr")
library("eulerr")
library("RColorBrewer")
library("e1071")
library("plotly") 
library("pheatmap")
library("msigdbr")
library("mitch")
library("rtracklayer")
library("dplyr")
library("org.Mm.eg.db")
library("fgsea")
  
})

CORES=16

```

# Sample manifest and alignment summary

Kallisto alignments for each sample was run against the cDNA + ncRNA index and compared the pseudoalignment rates were checked. The runs using the cDNA + ncRNA index showed higher alignment percentages. Hence, cDNA + ncRNA index was used for all downstream analyses.

```{r alignment, fig.width=11, fig.height=9, message=FALSE, warning=FALSE, echo=TRUE}

base<- "/mnt/vol1/Mouse_model_RNA_Seq/TAC/temp/"
idx<- "/mnt/vol1/Mouse_model_RNA_Seq/index"

samples<- read_tsv(file.path(base,"samples.txt"),col_names = "sample", show_col_types = FALSE)
samples$path <- file.path(base, samples$sample, "abundance.tsv")

align<- read_delim("/mnt/vol1/Mouse_model_RNA_Seq/TAC/alignment_summary.tsv")

df<- align%>% mutate(index = "cDNA + ncRNA", rate = 100 * pseudoaligned/processed) %>%
 select(sample, index, rate)


delta<- df %>% pivot_wider(names_from = index, values_from = rate) %>%
  mutate(delta = `cDNA + ncRNA`) %>%
  arrange(delta)
df$sample <- factor(df$sample, levels = delta$sample)

ggplot(df, aes(sample, rate, fill = index)) +
  geom_col(position = position_dodge(width = 0.75), width = 0.7) +
  labs(title = "Pseudoalignment rate by sample",
       x = "Sample", y = "Rate (%)", fill = "Run") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

```

# Import transcripts to gene level via trimport package

```{r message=FALSE, warning=FALSE, echo=TRUE}

tx2gene<- read.csv(file.path("/mnt/vol1/Mouse_model_RNA_Seq/tx2gene_r115.csv"), stringsAsFactors = FALSE)
stopifnot(all(c("TXNAME","GENEID") %in% names(tx2gene)))

files<- file.path(base, samples$sample, "abundance.tsv")
names(files) <- samples$sample
stopifnot(all(file.exists(files)))

txi<- tximport(files,
                type = "kallisto",
                tx2gene = tx2gene[, c("TXNAME","GENEID")],
                countsFromAbundance = "lengthScaledTPM",
                ignoreTxVersion = TRUE)

txi$counts[1:2, 1:6] 


out_dir <- "/mnt/vol1/Mouse_model_RNA_Seq/TAC"
dir.create(out_dir, showWarnings = FALSE, recursive = TRUE)

#counts matrix from tximport
txi_counts <- as.data.frame(txi$counts) %>%
  tibble::rownames_to_column(var = "ENSEMBL_GENEID") %>%
  dplyr::mutate(ENSEMBL_STRIP = sub("\\..*$", "", ENSEMBL_GENEID))

# map Ensembl to symbol (mouse)
txi_counts$SYMBOL <- AnnotationDbi::mapIds(
  org.Mm.eg.db,
  keys = txi_counts$ENSEMBL_STRIP,
  keytype = "ENSEMBL",
  column = "SYMBOL",
  multiVals = "first"
)

txi_counts <- txi_counts %>%
  dplyr::select(ENSEMBL_GENEID, SYMBOL, dplyr::everything(), -ENSEMBL_STRIP)

readr::write_csv(txi_counts, file.path(out_dir, "TAC_tximport_gene_counts_with_symbols.csv"))
saveRDS(txi, file.path(out_dir, "TAC_txi.rds"))


```

# Reading meta data that comes with the samples

```{r message=FALSE, warning=FALSE, echo=TRUE}

pheno<- read_excel("/mnt/vol1/Mouse_model_RNA_Seq/TAC/tac.xlsx",skip = 1)
colnames(pheno)

coldata<- pheno
rownames(coldata)<- coldata$ID

```

# Making DESeqDataSet from tximport ,filter the low counts and normalisation of the counts

DESeq2 object from the imported counts, set the design to use Group. Removed genes with very low total counts across samples (sum <= 10)

```{r message=FALSE, warning=FALSE, echo=TRUE}

dds<- DESeqDataSetFromTximport(txi, colData = coldata, design = ~ Surgery)

keep<- rowSums(counts(dds)) >= 10
dds<- dds[keep, ]

dds <- DESeq(dds)

norm_counts <- counts(dds, normalized = TRUE)

norm_df <- as.data.frame(norm_counts) %>%
  rownames_to_column("ENSEMBL_GENEID") %>%
  mutate(ENSEMBL_STRIP = sub("\\..*$", "", ENSEMBL_GENEID))

norm_df$SYMBOL <- AnnotationDbi::mapIds(
  org.Mm.eg.db,
  keys = norm_df$ENSEMBL_STRIP,
  keytype = "ENSEMBL",
  column = "SYMBOL",
  multiVals = "first"
)

norm_df <- norm_df %>%
  select(ENSEMBL_GENEID, SYMBOL, everything(), -ENSEMBL_STRIP)

write_csv(norm_df, file.path(out_dir, "TAC_DESeq2_normalized_counts_with_symbols.csv"))


levels(dds$Surgery)

```


# Unsupervised clustering analyses

I variance-stabilised the counts and used them to check sample structure. First, I plotted a sample–sample correlation heatmap (annotated Group) to spot outliers. Then I ran a PCA, colouring by Group, to see if groups separate. After that, fit the DESeq2 model with the interaction.

```{r PCA, fig.width=11, fig.height=9, message=FALSE, warning=FALSE, echo=TRUE}

vsd<- vst(dds, blind = TRUE)

ann<- as.data.frame(colData(dds))[, c("Surgery"), drop = FALSE]

pheatmap(cor(assay(vsd)),
         annotation_col = ann,
         annotation_row = ann) 

plotPCA(vsd, intgroup = c("Surgery"))


plotDispEsts(dds)

res <- results(dds,
               contrast = c("Surgery", "TAC", "Sham"),
               alpha = 0.05)

plotMA(res)

```


# Heatmap: all genes)


```{r heatmap_1, fig.width=11, fig.height=9, message=FALSE, warning=FALSE, echo=TRUE}


vsd_mat<- assay(vsd)
vsd_mat_scaled <- t(scale(t(vsd_mat)))


ens_full<- rownames(vsd_mat)
ens_stripped<- sub("\\.\\d+$", "", ens_full)
symbol_vec<- AnnotationDbi::mapIds(
  org.Mm.eg.db,
  keys     = ens_stripped,
  keytype  = "ENSEMBL",
  column   = "SYMBOL",
  multiVals = "first"
)

norm_expr_df<- as.data.frame(vsd_mat)
norm_expr_df$Gene<- ens_full       
norm_expr_df$Symbol <- unname(symbol_vec[ens_stripped])
norm_expr_df <- norm_expr_df[, c("Gene", "Symbol",
                                 setdiff(names(norm_expr_df),
                                         c("Gene", "Symbol")))]

write.csv(norm_expr_df,"/mnt/vol1/Mouse_model_RNA_Seq/TAC/normalised_vst_expression.csv", row.names = FALSE)

ann <- as.data.frame(colData(vsd))[, "Surgery", drop = FALSE]
ord_cols <- order(vsd$Surgery)

pheatmap(vsd_mat_scaled[, ord_cols],
         annotation_col = ann[ord_cols, , drop = FALSE],
         show_rownames = FALSE,
         cluster_cols  = TRUE,
         main = "All genes: TAC vs Sham")

```

# Heatmaps (top 40 genes by FDR)


```{r heatmap, fig.width=11, fig.height=9, message=FALSE, warning=FALSE, echo=TRUE}


top_heatmap_simple <- function(res, vsd, n = 40) {
  ord  <- order(res$padj, na.last = NA)
  keep <- ord[1:min(n, sum(!is.na(res$padj)))]
  
  genes <- rownames(res)[keep]
  mat   <- assay(vsd)[genes, , drop = FALSE]
  mat   <- t(scale(t(mat))) 
  
  ann <- as.data.frame(colData(vsd))[, "Surgery", drop = FALSE]
  
  pheatmap(mat,
           annotation_col = ann,
           show_rownames = TRUE,
           main = paste0("Top ", length(genes),
                         " DE genes: TAC vs Sham"))
}

## Top 40 DE genes
top_heatmap_simple(res, vsd, n = 40)


```

# Heatmaps of significant genes only


```{r het_sig, fig.width=11, fig.height=9, message=FALSE, warning=FALSE, echo=TRUE}

sig_idx <- which(res$padj <= 0.05 & !is.na(res$padj))
length(sig_idx) 

genes_sig <- rownames(res)[sig_idx]

mat_sig <- vsd_mat[genes_sig, , drop = FALSE]
mat_sig_scaled <- t(scale(t(mat_sig)))
  
ann <- as.data.frame(colData(vsd))[, "Surgery", drop = FALSE]
ord_cols <- order(vsd$Surgery)
  
pheatmap(mat_sig_scaled[, ord_cols],
           annotation_col = ann[ord_cols, , drop = FALSE],
           show_rownames = (length(sig_idx) <= 50),
           main = "FDR-significant genes (FDR <= 0.05): TAC vs Sham")


```


# Pathway enrichment Analysis


## GSEA (Hallmarks)

MSigDB “Hallmarks” (collection H) = 50 core biological with minimal overlap gene sets that capture broad, well-defined biological processes. 

```{r message=FALSE, warning=FALSE, echo=TRUE}

#Map ensembl
ens_all <- rownames(res)
ens_all_stripped <- sub("\\.\\d+$", "", ens_all)

ann_entrez <- tibble(ENSEMBL = ens_all_stripped) %>%
  distinct() %>%
  mutate(
    ENTREZID = AnnotationDbi::mapIds(
      org.Mm.eg.db,
      keys     = ENSEMBL,
      keytype  = "ENSEMBL",
      column   = "ENTREZID",
      multiVals = "first"
    )
  ) %>%
  filter(!is.na(ENTREZID))

#ranked stat vector (Wald statistic) on entrez iDs
rank_df <- as.data.frame(res) %>%
  rownames_to_column("ensgene") %>%
  mutate(ENSEMBL = sub("\\.\\d+$", "", ensgene)) %>%
  left_join(ann_entrez, by = "ENSEMBL") %>%
  filter(!is.na(stat), !is.na(ENTREZID))

rank_df_unique <- rank_df %>%
  group_by(ENTREZID) %>%
  summarise(stat = stat[which.max(abs(stat))], .groups = "drop")

ranks <- rank_df_unique$stat
names(ranks) <- rank_df_unique$ENTREZID
ranks <- sort(ranks, decreasing = TRUE)

#Hallmark gene sets (Mouse, Entrez IDs)
msig_entrez <- msigdbr(species = "Mus musculus", category = "H") %>%
  distinct(gs_name, entrez_gene) %>%
  group_by(gs_name) %>%
  summarise(genes = list(unique(entrez_gene)), .groups = "drop") %>%
  tibble::deframe()

# GSEA
gsea_hallmark <- fgseaMultilevel(
  pathways = msig_entrez,
  stats    = ranks,
  minSize  = 10,
) %>%
  arrange(padj) %>%
  mutate(contrast = "TAC vs Sham")

# top 20 Hallmark pathways
gsea_hallmark_20 <- gsea_hallmark %>%
  select(pathway, NES, padj, contrast) %>%
  head(20)


```

### Top significant GSEA (Hallmarks) pathways


I used Gene Ratio here. Gene ratio comes in clusterprofiler package under the dotplot.
In the gsea, it says 
"The leading-edge subset is a subset of genes within a gene set that drives the enrichment signal, while the "set size" is the total number of genes in the original gene set
"
So I used the \(\text{Gene Ratio}=\frac{\text{Count of  core enrichment genes}}{\text{Total count of genes in the pathway gene set}}\)  to calculate the Gene Ratio. 

References:https://github.com/YuLab-SMU/clusterProfiler/issues/364 



```{r Top_10_Hallmark, fig.width=11, fig.height=9, message=FALSE, warning=FALSE, echo=TRUE}

alpha <- 0.05

dot_df <- gsea_hallmark %>%
  filter(!is.na(padj), padj <= alpha) %>%
  mutate(
    direction = ifelse(NES > 0, "Enriched in TAC", "Enriched in Sham"),
    Count      = purrr::map_int(leadingEdge, length),
    GeneRatio  = Count / size,
    path_clean = gsub("^HALLMARK_", "", pathway),
    path_clean = gsub("_", " ", path_clean)
  ) %>%
  arrange(padj, dplyr::desc(abs(NES))) %>%
  group_by(direction) %>%
  slice_head(n = 10) %>%
  ungroup() %>%
  mutate(
    direction = factor(direction, levels = c("Enriched in TAC", "Enriched in Sham")),
    path_clean = factor(path_clean, levels = rev(unique(path_clean)))
  )

# DOT PLOT
ggplot(dot_df, aes(x = GeneRatio, y = path_clean)) +
  geom_point(aes(size = Count, colour = padj)) +
  facet_wrap(~ direction, ncol = 2, scales = "free_y") +
  scale_size_continuous(name = "Leading-edge genes") +
  scale_colour_gradient(low = "red", high = "blue", name = "FDR (padj)") +
  labs(
    title = "Top HALLMARK pathways in TAC vs Sham males",
    x = "GeneRatio (leading-edge / set size)",
    y = NULL
  ) +
  theme_bw(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 12),
    axis.text.y = element_text(face = "bold", size = 10),
    plot.title = element_text(face = "bold", hjust = 0.5),
    panel.grid.major = element_line(colour = "grey90", linewidth = 0.3)
  )

# BAR PLOT
bar_df <- dot_df %>%
  mutate(dir_simple = ifelse(NES > 0, "Enriched in TAC", "Enriched in Sham")) %>%
  arrange(NES)

bar_df$path_clean <- factor(bar_df$path_clean, levels = bar_df$path_clean)

ggplot(bar_df, aes(x = path_clean, y = NES, fill = dir_simple)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  labs(
    title = "Top HALLMARK pathways in TAC vs Sham males",
    x = NULL,
    y = "Normalised Enrichment Score (NES)",
    fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_text(face = "bold", size = 10),
    plot.title  = element_text(face = "bold", hjust = 0.5),
    legend.position = "left"
  )



```

## GSEA (GO:BP)

GO:BP (MSigDB C5, Biological Process) contains a larger set of more specific, ontology-based pathways that describe defined biological activities

```{r message=FALSE, warning=FALSE, echo=TRUE}

# GO: BP pathways (msigdbr), grouped by gs_name
msig_entrez <- msigdbr(species = "Mus musculus", category = "C5", subcategory = "GO:BP") %>%
  distinct(gs_name, entrez_gene) %>%
  group_by(gs_name) %>%
  summarise(genes = list(unique(entrez_gene)), .groups = "drop") %>%
  tibble::deframe()

# GSEA (TAC vs Sham)
gsea_GO_BP <- fgseaMultilevel(
  pathways = msig_entrez,
  stats    = ranks,
  minSize  = 10
) %>%
  arrange(padj) %>%
  mutate(contrast = "TAC vs Sham")

# top 20 GO_BP pathways
gsea_GO_BP_20 <- gsea_GO_BP %>%
  select(pathway, NES, padj, contrast) %>%
  slice_head(n = 20)

```

### Top significant GSEA (GO:BP) pathways

```{r Top_10_GO_BP, fig.width=11, fig.height=9, message=FALSE, warning=FALSE, echo=TRUE}


alpha <- 0.05

dot_df <- gsea_GO_BP %>%
  filter(!is.na(padj), padj <= alpha) %>%
  mutate(
    direction = ifelse(NES > 0, "Enriched in TAC", "Enriched in Sham"),
    Count      = purrr::map_int(leadingEdge, length),
    GeneRatio  = Count / size,
    path_clean = gsub("^GOBP_", "", pathway),
    path_clean = gsub("_", " ", path_clean)
  ) %>%
  arrange(padj, dplyr::desc(abs(NES))) %>%
  group_by(direction) %>%
  slice_head(n = 10) %>%
  ungroup() %>%
  mutate(
    direction  = factor(direction, levels = c("Enriched in TAC", "Enriched in Sham")),
    path_clean = factor(path_clean, levels = rev(unique(path_clean)))
  )



# BAR PLOT
bar_df <- dot_df %>%
  mutate(dir_simple = ifelse(NES > 0, "Enriched in TAC", "Enriched in Sham")) %>%
  arrange(NES)

bar_df$path_clean <- factor(bar_df$path_clean, levels = bar_df$path_clean)

ggplot(bar_df, aes(x = path_clean, y = NES, fill = dir_simple)) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey40") +
  labs(
    title = "Top GO:BP pathways in TAC vs Sham males",
    x = NULL,
    y = "Normalised Enrichment Score (NES)",
    fill = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.y = element_text(face = "bold", size = 10),
    plot.title  = element_text(face = "bold", hjust = 0.5),
    legend.position = "left"
  )


```


```{r fig.width=11, fig.height=9, message=FALSE, warning=FALSE, echo=TRUE}

sessionInfo()

```
